Episodic memory: mental time travel or a quantum `memory wave' function?

### 2019-01-10 boneyard.tex
backing up progress

>>>>

\textbf{KEY QUESTIONS}:
\begin{enumerate}
  \item Is this more of a ``review'' paper, or should I highlight the new analyses more?
  \item Who is this paper alienating and how can the narrative be reframed in a more inclusive way?
  \item Which ideas are ``new'' vs. restatements of old ideas?
  \item What are the key testable predictions to highlight, and how can they be communicated most clearly and effectively
  \item What's an appropriate venue for this?
\end{enumerate}

% Next to expand on:
% - Option 1: go into more depth about what the reinstatement functions look like for mental time travel vs. "quantum wave functions," under naturalistic vs. list-learning experiments.  Claim: different for naturalistic experiments but similar for list-learning experiments.
% - Option 2: get rid of the list learning stuff and transition directly to the "shape" of how experiences unfold.

\section*{How do our experiences and memories unfold over time?}
Although real-world experiences can exhibit rich temporal structures, it can be illustrative to consider the transformation of how we experience much simpler and less naturalistic events (e.g. studying random word lists) versus how we later remember those experiences.  Figure~\ref{fig:corrmats}A displays an average temporal correlation matrix describing the similarity structure of 1072 random word lists, each comprising 16 words~\citep{ZimaEtal18}.  I applied a topic model (trained on a curated corpus of Wikipedia articles) to yield a 100-dimensional embedding of each presented word~\citep{BleiEtal03}.  The left panel displays the average correlations between the topic vectors of each pair of presented words as a function of their presentation positions.  As shown, nearly all of the correlation ``mass'' is along the diagonal, and the off-diagonal entries are very close to zero.  Because the word orders were randomized, the temporal correlation matrix reflects that there are no consistent similarity patterns between any given pair of timepoints.  The right panel of Figure~\ref{fig:corrmats}A displays the average temporal correlation matrices describing participants' \textit{recalls} of those same lists.  Although the recall matrix exhibits the strongest correlations along its diagonal, some correlation mass bleeds into nearby recalls, reflecting (in part) participants' tendancies to recall semantically related words successively.\footnote{Note that participants recalled different numbers of words from each studied list; I used linear interpolation to resample each recall correlation matrix that preserved the size of the presentation correlation matrix.  This process resulted in some additional blurring along the diagonal.  In other words, the blurring along the diagonal reflects two phenomena: participants' tendancies to semanticly cluster their recalls, and an artifact of the temporal resampling.}  Differences in the temporal correlation structure between the original lists of words and the orders of participants' recalls illustrate one way that our memory systems can reorganize our experiences such that we recall them in a different way than we originally experienced them.

### 2019-01-10 main.tex
continuing work on paper, added some notes
backing up progress

It is difficult to explicitly test whether (apparent) semantic reinstatement in word list learning, as reflected in people's tendencies to semantically cluster recalls of random word lists, reflects integration of information across the experiences of studying those words.  However, remembering word lists does not \textit{require} integrating information across words in order to specifically gain a deep understanding of the list.  By design, there \textit{is} no deeper meaning of random word lists.

@@

\section*{BRAINSTORMING OUTLINE OF THE REST OF THE PAPER}
\begin{enumerate}
  \item \textit{Show video correlation matrix + recall correlation matrix.  Key idea: off-diagonal mass is introduced (primarily) by how people \textit{remember} the movie, not due to the contents of the movie itself (evidence: correlation matrices).  Further, this is not simply a matter of imprecision in how people recall, and/or our approach's ability to characterize recalled events (evidence: reinstatement figure showing which scenes are recalled similarly).  This sort of integration is how we ascribe meaning to events that unfold over disjointed extended timescales.}
  \item \textit{In theory this could also happen during \textit{any} sort of experience.  For example, when people study random word lists, they often re-order their recalls so that semantically related words are closer in the recall sequence than they were in the presented sequence.  This can be captured to an extent by semantic clustering measures.  However, it's difficult to examine this phenomenon in more depth because word lists necessarily lack the richness of more naturalistic stimuli.  There are no ``deeper ties'' between study events (i.e., individual words), so the sorts of integration operations that allow us to make sense of real-world experiences aren't given the same freedom of expression in list-learning experiments.  Evidence: presentation/recall correlation matrices during random free recall don't show the same sort of off-diagonal increase.}  Claim: mental time travel and quantum wave reinstatement look the same for random word list remembering, but very different for real-world remembering.
  \item The specific ties between unfolding events forms a scaffolding for understanding how those events are related.  This can be visualized geometrically by projecting word embedding representations of the events onto lower dimensional spaces (although see Flatland Fallacy paper, with HyperTools paper as a potential counterpoint).  This reveals another key property of how we remember: although there are substantial individual differences in the specific words people use to describe a shared experience, and how much detail they get into, the overall ``shapes'' of those recalled experience trajectories are highly similar across people.  This has potentially exciting implications for how we communicate memories to other people: we need to convey the ``gist'' (shape) of our experience.  It also has implications for how we learn by analogy: perhaps events that share schemas (cite Baldassano et al) unfold in similar ways (e.g. have similar shapes).  Learning which shapes go with which schemas could provide a scaffolding for rapid learning and generalization of new experiences, to the extent that those new experiences match a previously learned schema (shape).
    \item These ideas also have implications for other forms of learning (e.g. spatial learning) and neural decoding.  When we attempt to characterize thoughts (e.g. training a classifier to distinguish between states), we tend to train decoders to estimate the ``one'' state that the brain is in.  However, perhaps we should instead consider that our brains are contemplating, or holding in our thoughts, many possible states simultaneously.  Therefore we should consider the full distribution of states (rather than only considering its maximum) as a primary metric of characterizing mental function.  (Probably people have proposed this-- e.g. representing the posterior and using it to compute...need citations.)
    \end{enumerate}

\textbf{KEY QUESTIONS}:
\begin{enumerate}
  \item Is this more of a ``review'' paper, or should I highlight the new analyses more?
  \item Who is this paper alienating and how can the narrative be reframed in a more inclusive way?
  \item Which ideas are ``new'' vs. restatements of old ideas?
  \item What are the key testable predictions to highlight, and how can they be communicated most clearly and effectively
  \item What's an appropriate venue for this?
\end{enumerate}

\section*{How do our experiences and memories unfold over time?}
Although real-world experiences can exhibit rich temporal structures, it can be illustrative to consider the transformation of how we experience much simpler and less naturalistic events (e.g. studying random word lists) versus how we later remember those experiences.  Figure~\ref{fig:corrmats}A displays an average temporal correlation matrix describing the similarity structure of 1072 random word lists, each comprising 16 words~\citep{ZimaEtal18}.  I applied a topic model (trained on a curated corpus of Wikipedia articles) to yield a 100-dimensional embedding of each presented word~\citep{BleiEtal03}.  The left panel displays the average correlations between the topic vectors of each pair of presented words as a function of their presentation positions.  As shown, nearly all of the correlation ``mass'' is along the diagonal, and the off-diagonal entries are very close to zero.  Because the word orders were randomized, the temporal correlation matrix reflects that there are no consistent similarity patterns between any given pair of timepoints.  The right panel of Figure~\ref{fig:corrmats}A displays the average temporal correlation matrices describing participants' \textit{recalls} of those same lists.  Although the recall matrix exhibits the strongest correlations along its diagonal, some correlation mass bleeds into nearby recalls, reflecting (in part) participants' tendancies to recall semantically related words successively.\footnote{Note that participants recalled different numbers of words from each studied list; I used linear interpolation to resample each recall correlation matrix that preserved the size of the presentation correlation matrix.  This process resulted in some additional blurring along the diagonal.  In other words, the blurring along the diagonal reflects two phenomena: participants' tendancies to semanticly cluster their recalls, and an artifact of the temporal resampling.}  Differences in the temporal correlation structure between the original lists of words and the orders of participants' recalls illustrate one way that our memory systems can reorganize our experiences such that we recall them in a different way than we originally experienced them.

>>>>

It is difficult to explicitly test whether (apparent) semantic reinstatement in word list learning, as reflected in people's tendencies to semantically cluster recalls of random word lists, reflects integration of information across the experiences of studying those words.  However, remembering word lists does not \textit{require} integrating information across words in order to specifically gain a deep understanding of the list.  By design, there \textit{is} no deeper meaning of random word lists.  One way to characterize the meaning reflected by how an experience unfolds in time is through models that describe the geometric path that the experience takes through an appropriate representational space.

The specific ties between unfolding events forms a scaffolding for understanding how those events are related.  In addition to serving as a framework for quantifying the dynamic content of experiences, these trajectories may be visualized geometrically by projecting word embedding representations of the events onto lower dimensional spaces~\citep[e.g., ][]{HeusEtal18a}; for additional discussion and counterpoints see~\citep{JollChan18}.  \cite{HeusEtal18c} applied this approach to the same dataset reflected in Figures~\ref{fig:corrmats}, \ref{fig:reinstatement}, and~\ref{fig:trajectories}, reveals another key property of how we remember.  Although there are substantial individual differences in the specific words people use to describe a shared experience, and how much detail they get into, the overall ``shapes'' of those recalled experience trajectories are highly similar across people.  This has potentially exciting implications for how we communicate memories to other people: we need to convey the ``gist'' (shape) of our experience.  It also has implications for how we learn by analogy: perhaps events that share schemas~\citep[e.g., ][]{BaldEtal18}  unfold in similar ways (e.g., have similar shapes).  Learning which shapes go with which schemas could provide a scaffolding for rapid learning and generalization of new experiences, to the extent that those new experiences match a previously learned schema (shape).

% analysis idea: matrix of reinstatement functions for each timepoint-- e.g. considering each timepoint in turn (rows), how much are participants (on average) using language that mirrors how they talk about each other timepoint (columns).  there should be a strong diagonal, with some blockiness, and possibly some disjoint groups showing how events cluster.

@@

\section*{BRAINSTORMING OUTLINE OF THE REST OF THE PAPER}
\begin{enumerate}
  \item \textit{Show video correlation matrix + recall correlation matrix.  Key idea: off-diagonal mass is introduced (primarily) by how people \textit{remember} the movie, not due to the contents of the movie itself (evidence: correlation matrices).  Further, this is not simply a matter of imprecision in how people recall, and/or our approach's ability to characterize recalled events (evidence: reinstatement figure showing which scenes are recalled similarly).  This sort of integration is how we ascribe meaning to events that unfold over disjointed extended timescales.}
  \item \textit{In theory this could also happen during \textit{any} sort of experience.  For example, when people study random word lists, they often re-order their recalls so that semantically related words are closer in the recall sequence than they were in the presented sequence.  This can be captured to an extent by semantic clustering measures.  However, it's difficult to examine this phenomenon in more depth because word lists necessarily lack the richness of more naturalistic stimuli.  There are no ``deeper ties'' between study events (i.e., individual words), so the sorts of integration operations that allow us to make sense of real-world experiences aren't given the same freedom of expression in list-learning experiments.  Evidence: presentation/recall correlation matrices during random free recall don't show the same sort of off-diagonal increase.  Claim: mental time travel and quantum wave reinstatement look the same for random word list remembering, but very different for real-world remembering.}
  \item \textit{The specific ties between unfolding events forms a scaffolding for understanding how those events are related.  This can be visualized geometrically by projecting word embedding representations of the events onto lower dimensional spaces (although see Flatland Fallacy paper, with HyperTools paper as a potential counterpoint).  This reveals another key property of how we remember: although there are substantial individual differences in the specific words people use to describe a shared experience, and how much detail they get into, the overall ``shapes'' of those recalled experience trajectories are highly similar across people.  This has potentially exciting implications for how we communicate memories to other people: we need to convey the ``gist'' (shape) of our experience.  It also has implications for how we learn by analogy: perhaps events that share schemas (cite Baldassano et al) unfold in similar ways (e.g. have similar shapes).  Learning which shapes go with which schemas could provide a scaffolding for rapid learning and generalization of new experiences, to the extent that those new experiences match a previously learned schema (shape).}
    \item These ideas also have implications for other forms of learning (e.g. spatial learning) and neural decoding.  When we attempt to characterize thoughts (e.g. training a classifier to distinguish between states), we tend to train decoders to estimate the ``one'' state that the brain is in.  However, perhaps we should instead consider that our brains are contemplating, or holding in our thoughts, many possible states simultaneously.  Therefore we should consider the full distribution of states (rather than only considering its maximum) as a primary metric of characterizing mental function.  (Probably people have proposed this-- e.g. representing the posterior and using it to compute...need citations.)
    \end{enumerate}
